Approximation as Differentiation

### 2025-07-03 main/first-order.tex
Fix confusing overloading of C in 3.2.

We now consider how to adapt the CHAD framework~\cite{vákár22,nunes2023} to interpret programs as functions
along with their forward and backward approximation maps. Since the appropriate lattice of approximation
typically depends on the value being approximated, we will want to interpret a type $\tau$ as a pair $(X,
\partial X)$ where $X$ provides the conventional interpretation as a set and $\partial X$ provides an
appropriate lattice of approximations $\partial X(x)$ for every point $x \in X$. An expression $e$ of type
$\sigma$ with a single free variable of type $\tau$ (supposing $\tau$ and $\sigma$ denote $X$ and $Y$ in the
conventional semantics) must then be interpreted by a pair $(f, \partial f)$, with $f$ providing the
conventional interpretation as a function $X \to Y$, and $\partial f$ as a family of Galois connections
$\partial f(x): \partial X(x) \to \partial Y(f(x))$ taking approximants to approximants.

@@

In its general form, the Grothendieck construction $\Grothendieck{}T$ for an arbitrary indexed category $T:
\cat{C}^\op \to \Cat$, sometimes called the \emph{total category} for $T$, incorporates all the categories
$T(X)$ (for objects $X$ in $\cat{C}$) into a single category, together with morphisms that account for both
internal structure and reindexing along morphisms $f: X \to Y$ in $\cat{C}$.

\begin{definition}
\label{def:Grothendieck}
Suppose $T: \cat{C}^\op \to \Cat$ an indexed category. The \emph{Grothendieck construction}
$\Grothendieck{X}T(X)$ (also written $\Grothendieck{}T$) is the category which has as objects, all pairs $(X,
\partial X)$ of an object $X$ of $\cat{C}$ and an object $\partial X$ of $T(X)$, and as morphisms $(X,
\partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of morphisms $f: X \to Y$ in $\cat{C}$ and
morphisms $\partial f: \partial X \to T(f)(\partial Y)$ in $T(X)$.
\end{definition}

@@

For CHAD we are interested in the specific case of the Grothendieck construction where $\cat{C} = \Set$ and
each category $T(X) = \Func{X}{\cat{C}}$, the functor category which has as objects all $X$-indexed families
of objects of $\cat{C}$, and as morphisms $F \to G$, all families of morphisms $\eta(x): F(x) \to G(x)$ in
$\cat{C}$ (with naturality trivial because $X$ has only identity morphisms).

>>>>

We now consider how to adapt the CHAD framework~\cite{vákár22,nunes2023} to interpret programs as functions
along with their forward and backward approximation maps. Since the appropriate lattice of approximation
typically depends on the value being approximated, we will want to interpret a type $\tau$ as a pair $(X,
\partial X)$ where $X$ provides the conventional interpretation as a set and $\partial X$ provides an
appropriate lattice of approximations $\partial X(x)$ for every point $x \in X$. An expression $e$ of type
$\sigma$ with a single free variable of type $\tau$ (supposing $\tau$ and $\sigma$ denote $X$ and $Y$ in the
conventional semantics) must then be interpreted by a pair $(f, \partial f)$, with $f$ providing the
conventional interpretation as a function $X \to Y$, and $\partial f$ as a family of Galois connections
$\partial f(x): \partial X(x) \to \partial Y(f(x))$ taking approximations of $x$ to approximations of $f(x)$.

@@

In its general form, the Grothendieck construction $\Grothendieck{}T$ for an arbitrary indexed category $T:
\cat{D}^\op \to \Cat$, sometimes called the \emph{total category} for $T$, incorporates all the categories
$T(X)$ (for objects $X$ in $\cat{D}$) into a single category, together with morphisms that account for both
internal structure and reindexing along morphisms $f: X \to Y$ in $\cat{D}$.

\begin{definition}
\label{def:Grothendieck}
Suppose $T: \cat{D}^\op \to \Cat$ an indexed category. The \emph{Grothendieck construction}
$\Grothendieck{X}T(X)$ (also written $\Grothendieck{}T$) is the category which has as objects, all pairs $(X,
\partial X)$ of an object $X$ of $\cat{D}$ and an object $\partial X$ of $T(X)$, and as morphisms $(X,
\partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of morphisms $f: X \to Y$ in $\cat{D}$ and
morphisms $\partial f: \partial X \to T(f)(\partial Y)$ in $T(X)$.
\end{definition}

@@

For CHAD we are interested in the specific case of the Grothendieck construction where $\cat{D} = \Set$ and
the indexed category $T$ sends every set $X$ to the (functor) category $\Func{X}{\cat{C}}$ which has as
objects all $X$-indexed families of objects of a fixed category $\cat{C}$, and as morphisms $F \to G$, all
families of morphisms $\eta(x): F(x) \to G(x)$ in $\cat{C}$ (with naturality trivial because $X$ has only
identity morphisms).
